Analyzing Information Transfer from Different Data Sources via Precise Bias-variance Asymptotics

### 2021-07-06 revision_journal/results.tex
Updates from Overleaf

Equation \eqref{lem_cov_shift_eq} thus characterizes the variance of the HPS estimator, up to an error term on the order of $O\big(\frac{\sigma^2}{p^{-1/2}}\big)$ (recall that $\varphi > 4$, $c$ is an arbitrarily small constant, $\frac{p}{n_1+n_2} \in (2 {\tau}, \frac 2 {\tau})$).
Our result can be viewed as a generalization of the classical result in Lemma \ref{fact_tr} to the \textit{covariate shift} setting.
In particular, Lemma \ref{fact_tr} corresponds to a setting where the singular values $\lambda_1,\dots,\lambda_p$ are all equal to one, and $a = 1$. Then, the second part of equation \eqref{eq_a12extra} simplifies to
\[ \alpha_1 + \frac{p}{n_1 + n_2}\cdot \frac{\alpha_1}{\alpha_1 + \alpha_2} = \frac{n_1}{n_1 + n_2}. \]
By solving the above together with the first part of equation \eqref{eq_a12extra}, we obtain that
$$\alpha_1 = \frac{n_1}{n_1+n_2},\quad \alpha_2 = \frac{n_2-p}{n_1+ n_2}.$$
Plugging the above solutions back to the variance limit, we thus find that the limit is equal to
\begin{align*}
    \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}} &= \frac{\sigma^2}{n_1 + n_2} \frac{p}{\alpha_1 + \alpha_2} \tag{since the singular values of $M$ are equal to $1$} \\
    &= \frac{\sigma^2}{n_1 + n_2} \cdot\frac{p (n_1 + n_2)}{n_1 + n_2 - p} \tag{applying $\alpha_1$ and $\alpha_2$ from above} \\
    &= \frac{\sigma^2 p}{n_1 + n_2 - p},
\end{align*}
which is precisely the limit in Lemma \ref{fact_tr}.
Additionally, the convergence rate in equation \eqref{lem_cov_shift_eq} matches the one in Lemma \ref{fact_tr} for this special case.
%With Assumption \ref{assm_big1},
%$$ \frac{\sigma^2}{n_1+n_2}\bigtr{  \frac{1}{a_1 M(a)^\top M(a) + a_2  }  } \sim \frac{p \sigma^2}{n_1+ n_2}.$$
%Hence the right-hand side of \eqref{lem_cov_shift_eq} is much smaller than this main term by a factor of $p^{-1/2} (n_1+n_2)^{-1/2+2/\varphi + c}$, which we believe to be sharp up to the $(n_1+n_2)^c$ factor. Lemma \ref{fact_tr} can be also regarded as a special case of Theorem \ref{thm_main_RMT}. %also extends  to the inverse of the sum of two sample covariance matrices.
%To see this, when $M(a)=0$,
The proof of Theorem \ref{thm_main_RMT}, which is based on recent developments in random matrix theory \cite{Anisotropic}, can be found in Section \ref{appendix RMT}.
%For general $A_1,A_2$ that are not equal to one, we can still apply our result by rescaling $X^{(1)}$ and $M$ with $A_1 / A_2$.
%We defer a proof sketch of Theorem \ref{thm_main_RMT} until the end of the section.
%This amounts to replacing $M$ with $\frac{A_1}{A_2}M$ in Theorem \ref{thm_main_RMT}.

@@

%\paragraph{Illustrative examples.}
%Hence we take $a=1$ in \eqref{lem_cov_shift_eq}, and study the asymptotic limit
%$$\ell_{\var}(M):= \frac{\sigma^2}{n_1+n_2}\bigtr{  \frac{1}{a_1(M) \cdot M^\top M + a_2(M)  }  }, $$
%where $ M=(\Sigma^{(1)})^{1/2}(\Sigma^{(2)})^{-1/2}$, and $a_1(M)$ and $a_2(M)$ are defined through \eqref{eq_a12extra}.
\begin{example}[Covariate shift hurts more under imbalanced sample sizes]
Our second example illustrates that the effect of covariate shift worsens as the sample size of task one increases.
We consider a set of covariate shift matrices $M^{\top} M$ whose determinant are all equal to one, and the eigenvalues are bounded between $\tau^2$ and $1 / \tau^2$.
Let $\cS$ denote the set of $M\in\real^{p\times p}$ such that $M^{\top} M$ satisfies both conditions.
%\begin{align*}
%		\cS_{\mu}\define\bigset{M \left| \prod_{i=1}^p \lambda_i = \mu^p, \tau \le \lambda_p \le \lambda_1 \le \tau^{-1}\right.}.
%\end{align*}
\begin{claim}\label{prop_covariate}
%	Assume that $\Psi(\beta_1, \beta_2) = 0$ and $\rho_1, \rho_2>1$.
%	Let $g(M)$ denote the prediction loss of $\hat{\beta}_t^{\MTL}$ when $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} \in\cS_{\mu}$.
    Recall that $g(M) = \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}}$.
	We have the following relation between $g(\id_{p \times p})$ and the rest of $g(M)$ for any $M \in\cS$:
	\[ g(\id_{p\times p}) \le \bigbrace{1+ \bigo{\frac{n_2}{n_1}  }} g(M), \text{ for any } M \in \cS. \]
\end{claim}
\begin{proof}
    From equation \eqref{eq_a12extra}, it is easy to obtain the following estimates on $\alpha_1$ and $\alpha_2$ for any $M\in \cS$:
    \begin{align*}%\label{est_a12extra}
        \frac{n_1 - p}{n_1 + n_2} < \alpha_1 <  \frac{n_1 + n_2 - p}{n_1 + n_2},\quad \alpha_2 < \frac{n_2}{n_1 + n_2}.
    \end{align*}
    Applying the above into $g(M)$, and using the fact that the minimum eigenvalue of $ M^\top M$ is at least $\tau^2$, we obtain
    \begin{align}\label{approximateteM}
    \left(1+\frac{\rho_2}{(\rho_1-1)\mu_{\min}^2}\right)^{-1}h(M_0) \cdot \left(1 - \OO(p^{-\e})\right) \le g(M_0) \le h(M_0) \cdot \left(1 +\OO(p^{-\e})\right),
    \end{align}
    where
    $$h(M_0):=\frac{\sigma^2}{(\rho_1+\rho_2)a_1(M_0)}\cdot \frac1p\tr\left( \frac{1}{M_0^\top M_0}\right) .$$
    %With these two bounds, we can easily conclude \eqref{approxteM}. 
    %
    %We have that the test error satisfies
    %\be\label{approxteM}  te(M)\left(1 -  \frac{n_2}{n_1-p} \frac{1}{\lambda_p^2 + \frac{n_2}{n_1-p}}\right)  \le  \frac{\sigma^2}{n_1+n_2}\tr\left( \frac{1}{a_1M^\top M + a_2}\right) \le te(M),\ee
    %where $\lambda_p$ is the smallest singular value of $p$ and
    %$$te(M):= \frac{\sigma^2}{a_1(n_1+n_2)}\tr\left( \frac{1}{M^\top M}\right) .$$
    %Moreover, for all $M$ satisfying \eqref{GMcons}, the minimum of $te(M)$ is attained when $M= a\id$.
    By AM-GM inequality, we observe that 
    $$\tr\left( \frac{1}{M^\top M}\right) = \sum_{i=1}^p\frac{1}{\lambda_i^2}$$
    is minimized when $\lambda_1 = \cdots\lambda_p=\mu$ under the restriction $\prod_{i=1}^p\lambda_i\le \mu^p$. Hence we get that 
    \be\label{AMGM} h(M_0) \le \frac{\sigma^2}{\mu^2 (\rho_1+\rho_2)a_1(M_0)}.\ee

    On the other hand, when $M=\mu \id$, applying Lemma \ref{lem_cov_shift} we obtain that with high probability,
    \begin{align}\label{gvar_extra2}
    \begin{split}
    g(\mu \id)&= \frac{\sigma^2}{\rho_1+\rho_2}\cdot \frac1p\tr\left( \frac{1}{\mu^2 a_1 (\mu\id) + a_2(\mu\id)}\right) \cdot \left(1 +\OO(p^{-\e})\right)\\
    &\le \frac{\sigma^2}{\mu^2(\rho_1+\rho_2)a_1 (\mu\id)}.
    \end{split}
    \end{align}
    Combining \eqref{est_a12extra}, \eqref{approximateteM}, \eqref{AMGM} and \eqref{gvar_extra2}, we conclude the proof.
    %, we conclude that the sum $\sum_{i=1}^p\lambda_i^{-1}$ is smallest when $\lambda_1=\cdots=\lambda_p = a$.
\end{proof}
\todo{draw a connection to \citet{david2010impossibility}}
\end{example}

>>>>

Equation \eqref{lem_cov_shift_eq} thus characterizes the variance of the HPS estimator, up to an error term on the order of $O\big(\frac{\sigma^2}{p^{-1/2}}\big)$ (recall that $\varphi > 4$, $c$ is an arbitrarily small constant, $\frac{p}{n_1+n_2} \in (2 {\tau}, \frac 2 {\tau})$).
Our result can be viewed as a generalization of the classical result in Lemma \ref{fact_tr} to the \textit{covariate shift} setting.
In particular, Lemma \ref{fact_tr} corresponds to a setting where the singular values $\lambda_1,\dots,\lambda_p$ are all equal to one, and $a = 1$. Then, the second part of equation \eqref{eq_a12extra} simplifies to
\[ \alpha_1 + \frac{p}{n_1 + n_2}\cdot \frac{\alpha_1}{\alpha_1 + \alpha_2} = \frac{n_1}{n_1 + n_2}. \]
By solving the above together with the first part of equation \eqref{eq_a12extra}, we obtain that
$$\alpha_1 = \frac{n_1}{n_1+n_2},\quad \alpha_2 = \frac{n_2-p}{n_1+ n_2}.$$
Plugging the above solutions back to the variance limit, we thus find that the limit is equal to
\begin{align}
    \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}} &= \frac{\sigma^2}{n_1 + n_2} \frac{p}{\alpha_1 + \alpha_2} \tag{since the singular values of $M$ are equal to $1$} \nonumber \\
    &= \frac{\sigma^2}{n_1 + n_2} \cdot\frac{p (n_1 + n_2)}{n_1 + n_2 - p} \tag{applying $\alpha_1$ and $\alpha_2$ from above} \nonumber \\
    &= \frac{\sigma^2 p}{n_1 + n_2 - p}, \label{g_id}
\end{align}
which is precisely the limit in Lemma \ref{fact_tr}.
Additionally, the convergence rate in equation \eqref{lem_cov_shift_eq} matches the one in Lemma \ref{fact_tr} for this special case.
%With Assumption \ref{assm_big1},
%$$ \frac{\sigma^2}{n_1+n_2}\bigtr{  \frac{1}{a_1 M(a)^\top M(a) + a_2  }  } \sim \frac{p \sigma^2}{n_1+ n_2}.$$
%Hence the right-hand side of \eqref{lem_cov_shift_eq} is much smaller than this main term by a factor of $p^{-1/2} (n_1+n_2)^{-1/2+2/\varphi + c}$, which we believe to be sharp up to the $(n_1+n_2)^c$ factor. Lemma \ref{fact_tr} can be also regarded as a special case of Theorem \ref{thm_main_RMT}. %also extends  to the inverse of the sum of two sample covariance matrices.
%To see this, when $M(a)=0$,
The proof of Theorem \ref{thm_main_RMT}, which is based on recent developments in random matrix theory \cite{Anisotropic}, can be found in Section \ref{appendix RMT}.
%For general $A_1,A_2$ that are not equal to one, we can still apply our result by rescaling $X^{(1)}$ and $M$ with $A_1 / A_2$.
%We defer a proof sketch of Theorem \ref{thm_main_RMT} until the end of the section.
%This amounts to replacing $M$ with $\frac{A_1}{A_2}M$ in Theorem \ref{thm_main_RMT}.

@@

%\paragraph{Illustrative examples.}
%Hence we take $a=1$ in \eqref{lem_cov_shift_eq}, and study the asymptotic limit
%$$\ell_{\var}(M):= \frac{\sigma^2}{n_1+n_2}\bigtr{  \frac{1}{a_1(M) \cdot M^\top M + a_2(M)  }  }, $$
%where $ M=(\Sigma^{(1)})^{1/2}(\Sigma^{(2)})^{-1/2}$, and $a_1(M)$ and $a_2(M)$ are defined through \eqref{eq_a12extra}.
\begin{example}
Our second example illustrates that the effect of covariate shift worsens as the sample size of task one increases.
We consider a set of covariate shift matrices $M^{\top} M$ whose determinant are all equal to one, and the eigenvalues are bounded between $\tau^2$ and $1 / \tau^2$.
Let $\cS$ denote the set of $M\in\real^{p\times p}$ such that $M^{\top} M$ satisfies both conditions.
%\begin{align*}
%		\cS_{\mu}\define\bigset{M \left| \prod_{i=1}^p \lambda_i = \mu^p, \tau \le \lambda_p \le \lambda_1 \le \tau^{-1}\right.}.
%\end{align*}
\begin{claim}[Identical covariance provides approximately optimal transfer under imbalanced dataset sizes]\label{prop_covariate}
%	Assume that $\Psi(\beta_1, \beta_2) = 0$ and $\rho_1, \rho_2>1$.
%	Let $g(M)$ denote the prediction loss of $\hat{\beta}_t^{\MTL}$ when $M = \Sigma_1^{1/2}\Sigma_2^{-1/2} \in\cS_{\mu}$.
    Recall that $g(M) = \frac{\sigma^2}{n_1 + n_2} \bigtr{\frac 1 {\alpha_1 M^{\top} M + \alpha_2}}$.
	We have the following relation between $g(\id_{p \times p})$ and the rest of $g(M)$ for any $M \in\cS$:
	\begin{align} g(\id_{p\times p}) \le \bigbrace{1+ {\frac{n_2}{\tau^2\cdot n_1}  }} g(M), \text{ for any } M \in \cS. \label{eq_claim_id}
	\end{align}
\end{claim}
\begin{proof}
    We can write down the trace of $(\alpha_1 M^{\top} M + \alpha_2)^{-1}$ using the eigenvalues of $M^{\top} M$ as follows:
    \begin{align*}
        (\alpha_1 M^{\top} M + \alpha_2)^{-1} &= \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2} \\
        &\ge \sum_{i=1}^p \frac{1} {\alpha_1 \lambda_i^2 + \alpha_2 \cdot {\lambda_i^2} / {\tau^2}} \tag{since $\lambda_i \le \tau$, for any $i$} \\
        &= \frac 1 {\alpha_1 + \alpha_2 / \tau^2} \sum_{i=1}^ p \frac 1 {\lambda_i^2} \\
        &\ge \frac 1 {\alpha_1 + \alpha_2 /\tau^2} p\cdot \bigbrace{\prod_{i=1}^p \frac 1 {\lambda_i^2}}^{1/ p} \tag{by the AM-GM inequality} \\
        &= \frac{p}{\alpha_1 + \alpha_2 / \tau^2} \tag{since $\prod_{i=1}^p \lambda_i^2 = \det(M^{\top} M) = 1$}
    \end{align*}
    Next, recall from equation \eqref{g_id} that
    $g(\id_{p\times p}) = \frac{\sigma^2 p}{n_1 + n_2 - p}$.
    Thus, equation \eqref{eq_claim_id} follows if we show that (after rearranging terms)
    \begin{align}
        \frac{n_1 + n_2}{n_1 + n_2 - p} \cdot \frac 1 {\alpha_1 + \frac{\alpha_2}{\tau^2}} \le 1 + \frac{n_2}{\tau^2 \cdot n_1}. \label{eq_claim_id_step}
    \end{align}
    From the second part of equation \eqref{eq_a12extra}, we have that
    \begin{align*}
        \frac{n_1} {n_1 + n_2} &= \alpha_1 + \frac 1 {n_1 + n_2} \bigbrace{\sum_{i=1}^p \frac{\lambda_i^2\alpha_1}{\lambda_i^2\alpha_1 + \alpha_2}} \\
        &< \alpha_1 + \frac p {n_1 + n_2}.
    \end{align*}
    Thus, $\alpha_1 > \frac{n_1 - p}{n_1 + n_2}$, which implies that $\alpha_2 < \frac{n_2}{n_1 + n_2}$.
    Hence, $\alpha_1 + \frac{\alpha_2}{\tau^2} = 1 - \frac{p}{n_1 + n_2} - \alpha_2 + \frac{\alpha_2}{\tau^2} \le 1 - \frac{p}{n_1 + n_2} + \frac{n_2}{n_1 + n_2} \frac 1 {\tau^2}$, which implies
    equation \eqref{eq_claim_id_step} by straightforward calculations.
    Thus, the proof is complete.
\end{proof}
\todo{draw a connection to \citet{david2010impossibility}}
\end{example}
